\documentclass[12pt,reqno]{amsart}
\usepackage{mathpazo}
\usepackage[scaled=0.9]{helvet}

\newcommand{\newterm}[1]{\emph{#1}}

\title{Consumption: Certainty}
\author{Jyotirmoy Bhattacharya}
\address{Ambedkar University Delhi}
\email{jyotirmoy@jyotirmoy.net}
\date{\today}

\begin{document}
\maketitle
\section{Two-period case}
\subsection{Budget constraint}
Consider a consumer who lives for two periods, has an endowment of
$y_1$ and $y_2$ units of goods in the two periods respectively and can
borrow and lend any amount that they like at the real rate of interest
$r$. 

Suppose the consumer consumes $c_1$ in the
first period. Then she will have to take a loan of $c_1-y_1$ to
finance her consumption. (This number can be negative, in which case
the consumer is lending rather than borrowing.) In the next period the
consumer will therefore have to make loan repayments of
$(1+r)(c_1-y_1)$. Assume that the consumer does not want to make any
bequests and cannot die with any outstanding loans, consumption in the
second period must be,
\[c_2=y_2-(1+r)(c_1-y_1)\]
Simplifying and rearranging we have
\begin{equation}\label{eq:two-period-budget}
c_1+\frac{c_2}{1+r}=y_1+\frac{y_2}{1+r}
\end{equation}
This is the budget constraint faced by the consumer. We can interpret
this to mean that the present value of the consumer's consumption
stream must equal the present value of their incomes.

\subsection{Utility maximization}
Suppose the consumer maximises a quasiconcave utility function
$U(c_1,c_2)$ subject to this budget constraint. Then the consumer's
first-order conditions are
\begin{align}
U_1(c_1,c_2)&=\lambda\\
U_2(c_1,c_2)&=\lambda/(1+r)
\end{align}
where $\lambda$ is the Lagrange multiplier corresponding to the budget
constraint and $U_i(c_1,c_2)$ denotes the partial derivative $\partial
U/\partial c_i$. We have explicitly shown the dependence of the
partial derivatives on the value of consumption in both periods. These
first-order conditions along with the budget
constraint~\eqref{eq:two-period-budget} together determines the value
of $c_1$, $c_2$ and~$\lambda$.
\subsection{Comparative statics}
Assuming that consumption in both periods is a normal good, an
increase in either $y_1$ or $y_2$ increases both $c_1$ and $c_2$.

The effects of a change in $r$ are ambiguous. An increase in $r$ makes
consumption in period~$2$ relatively cheap compared to consumption in
period~$1$. Therefore the substitution effect causes $c_1$ to decrease
and $c_2$ to increase. It is traditional to decompose the income
effect into two parts. First, an increase in $r$ reduces the present
value of the consumer's endowments and hence decreases his real
income. Second, an increase in $r$, by making the consumption in
period~$2$ cheaper increases his real income.\footnote{
For more about the Slutsky equation in the case of a consumer with
fixed endowments of goods see section~9.1 in Varian's
\emph{Microeconomic Analysis}, 3rd ed.} The sign of the
resultant of these two effects on consumption depends on whether the
consumer is a net lender in period~1 and a net borrower in period~2 or
vice-versa. In case the consumer is a net lender in period~1 and a net
borrower in period~2 the net income effect is positive. Assuming the
consumption in both periods in a normal good, this means that the
substitution effect and the income effect act in opposite directions
on $c_1$ in this case leading to an ambiguous effect.

\section{Many periods}
Assume that rather than just living for two periods the consumer lives
for $T+1$ periods. Further assume that the real rate of interest takes a
constant value $r$ over the consumer's lifetime. For convenience we
define $\delta=1/(1+r)$. It is also convenient to start time from
period~0 rather than period~1.

\subsection{Budget constraint}
Arguing as before, the consumer's budget constraint is
\begin{equation}\label{eq:many-period-budget}
\sum_{i=0}^T \delta^i c_i = \sum_{i=0}^T \delta^i y_i
\end{equation}

\subsection{Utility function}
We could proceed as before by assuming a utility function
$U(c_0,\ldots,c_T)$ and deriving the first order conditions. However,
because the marginal utility in each period depends on consumption in
all periods it is hard to draw any sharp conclusions at this level of
generality. Therefore we need to impose some restrictions on the form
of the utility functions.

Suppose, for example we assume that the utility function is additively
separable, i.e.

\begin{equation}\label{eq:utility-addsep}
U(c_0,\ldots,c_T)=v_0(c_0)+v_1(c_1)+\cdots+v_T(c_T)
\end{equation}

Then the first-order conditions take the form

\begin{equation}\label{eq:foc-additively-separable}
v_i'(c_i) = \delta^i \lambda \qquad i=0,\ldots,T
\end{equation}
where, as before, $\lambda$ is the Lagrange multiplier corresponding
to the budget constraint.

Sometimes we want to restrict the consumers preferences even further,
by assuming that the different $v_i$ differ from each other by only a
geometric discounting factor.

\begin{equation}\label{eq:utility-geometric}
U(c_0,\ldots,c_T)=\sum_{i=0}^T \beta^i u(c_i)
\end{equation}
where $\beta$ is a constant, referred to as the subjective rate of
discount, such that $0<\beta<1$.  

In this case the first-order conditions take the particularly simple
form
\begin{equation}\label{eq:foc-stationary}
u'(c_i)=\left(\frac{\delta}{\beta}\right)^i \lambda \qquad i=0,\ldots,T
\end{equation}

In case $\delta=\beta$, this implies that $u'(c_i)$ is the same for
all $i$, which, assuming that $u'(\cdot)$ is a strictly decreasing
function, means that $c_i$ is constant for all $i$. The present
period's income does not influence the present period's consumption at
all. Consumption is determined solely by lifetime resources as given
by~\eqref{eq:many-period-budget}.

The case $\delta \neq \beta$ is also instructive. Suppose
$\delta>\beta$. In this case it follows from~\eqref{eq:foc-stationary}
that consumption decreases over time. Formally, this is because if
$\delta>\beta$ then by~\eqref{eq:foc-stationary} $u'(c_i)$ increases
over time, and since $u'(c)$ is a decreasing function of consumption,
this implies that $c$ decreases over time. 

The economic logic behind this result is that $\delta$ is the number of
units of consumption we have to give up at present in order to
purchase one more unit of consumption next period, whereas $\beta$ is
the number of units of marginal utility we are willing to give up at
present in order to have one more unit of marginal utility in the next
period. Suppose we start with the same consumption $c$ in this period and
the next. If we reduce consumption in the next period by a small
amount $\Delta c$ then at the prevailing market prices we can
increase present consumption by $\delta\Delta c$. The increase in
utility from the increase in present consumption is approximately
$u'(c)(\delta\Delta c)$.\footnote{We are using Taylor's theorem:
  $u(c+\delta\Delta c)-u(c) \approx u'(c)(\delta\Delta c)$} The decrease in utility from the reduction in
next period's consumption is approximately $\beta u'(c)(\Delta c)$. The
net change in utility would be $(\delta-\beta)u'(c)(\Delta c)$ which
is positive when $\delta>\beta$. Thus it is beneficial to increase present consumption and
reduce future consumption if we are starting from a position of
equality. Indeed, it will be optimal to increase consumption in the
present period (say period $i$) and decrease consumption in the next
period (period $i+1$) till the following equality between the MRS and
the price ratio is satisfied,
\[\frac{u'(c_{i+1})}{u'(c_i)}=\frac{\delta}{\beta}\]
\subsection{Exogenous variables}
It is possible to unify~\eqref{eq:utility-addsep}
and~\eqref{eq:utility-geometric} by writing
\[v_i(c_i)=\beta^i u(c_i,\xi_i)\]
where $\xi_i$ is an exogenous variable such a the consumer's age or
the number of members in the household. In this case the first-order
conditions become
\[u'(c_i,\xi_i)=\left(\frac{\delta}{\beta}\right)^i \lambda \qquad
i=0,\ldots,T\]
Knowing how $\xi$ affects the marginal utility would
now let us make some predictions regarding the path of consumption.

\subsection{Comparative statics}
Assuming that consumption in every period is a normal good, an
increase in $y_i$ increases every $c_i$.

The effect of an increase in $r$, or equivalently, a decrease in
$\delta$ remains ambiguous because of the same income and substitution
effects as discussed earlier. But for the utility function given
by~\eqref{eq:utility-geometric}, we can say a little
more. From~\eqref{eq:foc-stationary} we can see that a decrease in
$\delta$ means that the \emph{growth rate} of consumption slows
down. Remember that even in this case we do not have any information
regarding the \emph{level} of consumption in any period since the
level would depend on $\lambda$ which in turn depends on $\delta$.
\end{document}